Distributed Tic Tac Toe:\\ Peer-to-Peer Team-Based Gaming System

### 2017-12-01 paper/paper.tex
Fixed some glaring mistakes in the paper. Recompile before submitting

\begin{abstract}
This article describes a Distributed Multiplayer Tic-Tac-Toe software game written in Java using the RMI API. While previous implementations attempting to serve a similar purpose exist, none of them boast good design or efficiency, and neither of them deliver the same functionality. This project allows clients to play a game of Tic-Tac-Toe locally, with a twist: players collecively decide on what the next move should be by voting in teams. The implementation is derived from a design that follows the distributed applications guidelines and is prepared to handle corner cases such as client disconnects.
\end{abstract}

\section{Introduction}
Programming is very much an in-demand profession in the current job market, as well as being a popular hobby. Most beginners take their first steps in writing code
following a simple, single-threaded model with zero distribution. It is often difficult for new programmers to abstract the key knowledge they have obtained to a point where they can independently apply it to a different scenario.

Distributed software engineering principles are evolving quickly – the number of Internet-enabled devices has rapidly increased in recent years, and the level of complexity in creating programs to use these interconnected systems effectively has increased along with it. This complexity arises from core concepts such as co-ordinating resource sharing and fault tolerance - these are challenging concepts to learn for a novice programmer, but they are integral to creating a reliable and effective distributed system.

We believe that programs which serve as examples of turning familiar, relatively simple local applications into distributed ones are key to bridging this gap, providing a hands-on introduction to a field which is becoming ever more important with the deployment of novel concepts like fog computing. The sheer number of programming tutorials available on the internet shows that there is a demand for model solutions for educational purposes. Games in particular have proven to be an effective means of communicating programming methods and techniques in an engaging way \cite{Kiniry:2011:VG:1984674.1984681}. For this reason, we have decided to take a simple, well known game and develop it into a distributed system: Tic-Tac-Toe played as a peer-to-peer, team-based gaming system where the players are organised into teams and required to vote the moves their team will make.

@@

Upon initialising the game, the player will take a note of their address and port number. As the game is played on a local network, the player will attempt to connect to all other existing nodes on the network. If none are found, then the client will initialise a new game instance and set itself as one of the leaders.

If a player manages to connect to another machine on the local network (ie. that machine is paricipating in an active game session), they will retrieve the current game state, check if they are one of the leaders, print the current board state, and prepare for the turn.

The leaders update a common game state at the end of each event before broadcasting this updated game state to all team members. Players are told by the leaders when the game, and therefore the session, ends.

\section{Implementation}

The implementation follows the RMI distributed object model. A GameInt interface extends java.rmi.Remote and lists remote classes.

A Game class implements GameInt and extends the UnicastRemoteObject. It is an analogy for what a player knows and does during a game. It provides an implementation for all the listed classes in the interface which include the functionality of requesting the player to make a play, setting the leader of the game, several methods to interact with the Role of the player, and a main method that connects to other game instances.

Currently, when running the game, the user supplies a local instance name and the name of another instance as command line arguments. Each instance then uses this information to do an RMI Registry lookup and get a reference to the other client.

The Role interface allows the application to differentiate between Leader and Player classes which extend it. It utilises the State design pattern to allow players to seamlessly have their role changed when required (eg. if the current leader disconnects).

A Leader is responsible for collecting the votes from all of the members in their team in the turnStarts() method. This is done asynchronously using a separate thread for each member – an instance of VoteThread. Each thread collects a vote from one assigned player then writes it into a corresponding position in a shared AtomicIntegerArray that they have all been passed by the leader. This ensures the main leader thread can then read these values.

Each Leader then informs all players in the game about the most popular vote chosen by their team. In the event of two or more moves receiving an equal number of votes, a leader will randomly select  A Player instance represents a regular player that can only vote, it cannot independently generate and broadcast a new game state.

@@

Elections use the bully algorithm to determine which Player instance will become the new Leader instance for their team. The other team members and the leader of the opposing team are all informed of the new leader, creating an altered topology to compensate for the failure:

\begin{figure}[h]
	\includegraphics[width=\linewidth]{images/DAS-topology-election.png}
	\caption{Proposed Topology for UDP-based Version of System}
	\label{fig:topology-election}
\end{figure}

If the failed machine joins the game session again, it will not automatically regain Leader status, instead participating as a regular Player instance.

All plays are recorded on a GameState object instance which acts as a game board. After each turn, GameState checks if the game is over and relays this message to the leaders which can then inform the members of their team.

@@

\section{Related Work}
While other distributed multiplayer games already exist, they do not serve the same
purpose as our project. Existing products can be split into two categories: too simple and too complex. Examples of the former generally deal with the difficulties of distributed systems in a bad fashion, promoting questionable design and programming practices. One specific example of this is a YouTube guide to building a single player-per-team distributed tic-tac-toe game in Java \cite{MrWayFarOutYoutube15} - while the material is accessible for inexperienced networks programmers, it puts all functionality in a single class and does not comply with any design guidelines. As such, this resource has the potential to further confuse learners about programming distributed systems when they try to read more complex documentation which does obey software engineering design principles.

The latter category consists of resources which provide details of state-of-the-art practices and advanced optimisations. An example of this is a version of the game Quake II, modified to run multiplayer sessions more efficiently over a distributed system \cite{DBLP:conf/nsdi/BharambePS06}. This is excellent material for an experienced programmer looking to refine their skills, but daunting to a newcomer due to a high level of complexity in both networking concepts and game logic.

@@

The produced software serves as a good demonstration of a distributed game. The use of a simple, well known game turned into a distributed, multiplayer experience highlights the benefits and complexities associated with distributed systems, and allows the technology to be demonstrated in an accessible format, allowing users of all levels to make use of the software without being discouraged by an overwhelming level of complexity.

@@

In the project's current state, leaders are appointed using the bully algorithm when one machine acting as a leader fails. However, there is no support for failure of communication from both leaders, i.e. if both leaders were to crash at the same time. Although the likelihood of this occurring is slim, this possibility still needs to be addressed to improve player experience and to improve the standard of our product as an educational resource.

Another improvement which we would make in further iterations would be some form of automatic team balancing - currently, no replacement team members are allocated to an underpopulated team if several peers drop out of a session. Another situation which needs to be addressed is the scenario of all members of a team leaving a game before it is complete, an event which currently has no resolution.

>>>>

\begin{abstract}
This article describes a distributed multiplayer Tic-Tac-Toe game
written in Java using the RMI API. While previous implementations
attempting to serve a similar purpose exist, none of them boast good
design or efficiency, and neither of them deliver the same
functionality. Our implementation allows clients to play a game of
Tic-Tac-Toe with a twist: players collectively decide on what the next
move should be by voting in teams. The implementation is derived from
a design that follows the distributed applications guidelines and is
prepared to handle corner cases such as client disconnects.
\end{abstract}

\section{Introduction} % TODO: cite the first 2 paragraphs
Programming is very much an in-demand profession in the current job market, as well as being a popular hobby. Most beginners take their first steps in writing code
with no concurrency. It is often difficult for new programmers to abstract the key knowledge they have obtained to a point where they can independently apply it to a different scenario.

Distributed software engineering principles are evolving quickly --- the number of Internet-enabled devices has rapidly increased in recent years, and the level of complexity in creating programs to use these interconnected systems effectively has increased along with it. This complexity arises from core concepts such as co-ordinating resource sharing and fault tolerance --- these are challenging concepts to learn for a novice programmer, but they are integral to creating a reliable and effective distributed system.

We believe that programs which serve as examples of turning familiar, relatively simple local applications into distributed ones are key to bridging this gap, providing a hands-on introduction to a field which is becoming ever more important with the deployment of novel concepts like fog computing. The sheer number of programming tutorials available on the internet shows that there is a demand for model solutions for educational purposes. Games in particular have proven to be an effective means of communicating programming methods and techniques in an engaging way \cite{Kiniry:2011:VG:1984674.1984681}. For this reason, we have decided to take a simple, well known game and develop it into a distributed system: Tic-Tac-Toe played as a peer-to-peer, team-based gaming system, where the players are organised into teams and required to vote the moves their team will make.

@@

A new game instance tries to join an already existing game specified
by the server name and an RMI Registry handle. If a connection
succeeds (ie. that machine is paricipating in an active game session),
the new player will retrieve the current game state, initialise
differently depending on whether it is assigned to be a leader or not, print the current board state, and prepare for the turn.

The leaders update a common game state at the end of each event before
broadcasting this updated game state to all team members. Each player
checks if the received play ends the game. If so, it stops all
activity and, after the RMI connections close, exits.

\section{Implementation} % TODO: define RMI

The implementation follows the RMI distributed object model. A
\texttt{GameInt} interface extends \texttt{java.rmi.Remote} and lists
all methods that can be called using RMI.

A \texttt{Game} class implements \texttt{GameInt} and extends
\texttt{UnicastRemoteObject}. It is the main class for each player,
holding most of its knowledge and behaviour. It provides an
implementation for all the listed methods in the interface, which
include the functionality of requesting a player to make a play,
setting the leader of the game, several methods to access
role-specific functionality, and a main method that connects to other game instances.

Currently, when running a game, the user supplies a local instance
name and the name of another instance as command line arguments. Each
instance then uses this information to do an RMI Registry lookup and
get a reference to the other peer in an attempt to establish a
session. This allows multiple different sessions to run simultaneously
on the same network without interference --- ideal for educational use, where multiple groups are likely to be working on the same internal network.

% TODO: reference for state design pattern
The \texttt{Role} interface allows the application to differentiate between \texttt{Leader} and \texttt{Player} classes which extend it. It utilises the State design pattern to allow players to seamlessly have their role changed when required (eg. if the current leader disconnects).

A Leader is responsible for collecting votes from all of the members
in their team in the \texttt{turnStarts()} method. This is done
asynchronously using a separate thread for each member --- an instance
of \texttt{VoteThread}. Each thread collects a vote from one assigned
player and writes it into a corresponding position in a shared
\texttt{AtomicIntegerArray}. The leader then uses these votes to decide
on a play.

% TODO: stick to uppercase or lowercase leaders, players
Each \texttt{Leader} then informs all players in the game about the
most popular vote chosen by their team. In the event of two or more
moves receiving an equal number of votes, a leader will randomly
select one of them. A \texttt{Player} instance represents a regular
player that can only vote, it cannot independently generate and
broadcast a new game state.

@@

% TODO: reference the bully algorithm
Elections use the ``Bully'' algorithm to determine which \texttt{Player} instance will become the new \texttt{Leader} instance for their team. The other team members and the leader of the opposing team are all informed of the new leader, creating an altered topology to compensate for the failure:

\begin{figure}[h]
	\includegraphics[width=\linewidth]{images/DAS-topology-election.png}
	\caption{Proposed Topology for UDP-based Version of System} % TODO: is this caption out of date?
	\label{fig:topology-election}
\end{figure}

If the failed machine joins the game session again, it will not automatically regain \texttt{Leader} status, instead participating as a regular \texttt{Player} instance.

All plays are recorded in a \texttt{GameState} object instance, which
acts as a game board. After each turn, \texttt{GameState} checks if
the game is over. If it is, the client quietly waits for the RMI
connections to time out.

@@

\section{Related Work}
While other distributed multiplayer games already exist, they do not serve the same
purpose as our project. Existing products can be split into two categories: too simple and too complex. Examples of the former generally deal with the difficulties of distributed systems in a bad fashion, promoting questionable design and programming practices. One specific example of this is a YouTube guide to building a single player-per-team distributed tic-tac-toe game in Java \cite{MrWayFarOutYoutube15} --- while the material is accessible for inexperienced networks programmers, it puts all functionality in a single class and does not comply with any design guidelines. As such, this resource has the potential to further confuse learners about programming distributed systems when they try to read more complex documentation, which does obey software engineering design principles.

The latter category consists of resources that provide details of state-of-the-art practices and advanced optimisations. An example of this is a version of the game Quake II, modified to run multiplayer sessions more efficiently over a distributed system \cite{DBLP:conf/nsdi/BharambePS06}. This is excellent material for an experienced programmer looking to refine their skills, but daunting to a newcomer due to a high level of complexity in both networking concepts and game logic.

@@

The produced software serves as a good demonstration of a distributed game. The use of a simple, well-known game turned into a distributed, multiplayer experience highlights the benefits and complexities associated with distributed systems, and allows the technology to be demonstrated in an accessible format, allowing users of all levels to make use of the software without being discouraged by an overwhelming level of complexity.

@@

In the project's current state, leaders are appointed using the
``Bully'' algorithm when a leader fails. However, there is no support
for failure of communication from both leaders, i.e. if both leaders
were to crash at the same time. Although the likelihood of this
occurring is slim, this possibility still needs to be addressed to
improve player experience and to improve the standard of our product
as an educational resource.

Another improvement, which we would make in further iterations, would be some form of automatic team balancing --- currently, no replacement team members are allocated to an underpopulated team if several peers drop out of a session. Another situation which needs to be addressed is the scenario of all members of a team leaving a game before it is complete, an event which currently has no resolution.
